---
title: "groundhogR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{groundhogR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Front-end overview

The function `groundhog.library()` substitutes the functions: `library()` and
`install.packages()`.
  
In contrast to `library()`, with `groundhog.library()`

- A date must be included within the `groundhog.library()` command
- The version of the package available on that day, is automatically installed if needed
- All dependencies for that package are installed as available that day.
 
groundhogR allows having multiple (possibly all) versions of a package
available, and loading whichever one is wanted. For an older project you work
with an older version of a package to ensure reproducibility, while for a new
project you work with a newer version.

For example, to work with the package rio as it was available on 2017-09-15 
instead of 

```{r}
library("rio")
```

you run

```{r}
groundhog.library("rio", "2017-09-15")
```

Even on the same script (`.R` file) you can switch which version you use, like 
this:

```{r}
groundhog.library("rio", "2019-01-27")
```

# Recommended uses

## USE 1: Writing new R scripts

When working on new scripts, unless you have a reason not to, choose a date for
the whole script and stick to it. Maybe the date when the project was started.
It makes sense to set the date as a variable you can call though your script.
You can give that variable any name, perhaps `groundhog.day` makes sense.

So the top of your reproducible R scripts will look something like this:

```{r}
####   Written in R Version 3.6.1 - ####

library("groundhogR")

groundhog.date <- "2019-04-21"
groundhog.library("rio", date = groundhog.day)
groundhog.library("pwr", date = groundhog.day)

# <your code>
```

## USE 2: Revive existing scripts

The whole reason for groundhogR to exist is that packages change over time in
ways that make old scripts not work. A likely use of groundhogR is thus to
revive old scripts that no longer work.

To recover a non-working script replace its `library()` statements, with
`groundhog.library()` ones, trying older dates until the script works again.

You can start setting the date to the date when the script you are reviving was
last saved. If that does not work, roll back the clock further (and possibly
separately for each package in the script).

This process is facilitated by knowing when a package had a version change.
Use `groundhogR::toc()` to get the table of contents, like this: 

```{r}
groundhogR::toc("rio")
groundhogR::toc("pwr")
```

To get a single set with all relevant changes worth trying:

```{r}
groundhogR::cross.toc(c("rio", "pwr", "R"))
```
 
## USE 3: Document changes in packages 

To explain an error in a paper, or document the introduction of a new feature,
etc., it may be useful to run within the same script the same code for two
versions of the same package.

For example, one may run some commands with one version of a package and later
with a different version (useful for debugging and explaining errors caused by
package changes).

If an older version of a package is not compatible with a newer version of R,
the needed version of R will be indicated. The user can install it and run the
same script on it.

Note that RStudio can manage multiple versions of R and users select the desired
one when starting RStudio.

<pending>: want to create the command changeR() which will switch the version of 
R being used

# Back end

Two key design features that enable groundhogR to maintain multiple versions of
the same package within the same R installation, and calling specific versions
on the spot:

1. Richer directory structure for the library of packages, and 
1. Dynamically generated search path of packages modified with each 
`groundhog.library()` call.

## Directory structure.

With base R, there is a single library folder with package names (without
indicating version) as subfolders. Those subfolders contain only the latest
installed version of the corresponding package, and the files are deleted and
replaced when a package gets updated. [^1]

[^1]: It is possible to have multiple libraries with base R, and many users do
without awareness have more than one attached in `.libPaths()`. In practice,
users tend to install all packages in the same library. More importantly, the
name of subfolders in base R is just the name of the package, while in
groundhogR it is the name and version combined: `package_version`.

groundhogR, instead, gives a different subfolder to each package version. When a
new version is installed, it is stored in a new subfolder, keeping all existing
ones unchanged.

So for example, in base R, the package rio 0.5.1 is stored here

```
C:\Users\mike\Documents\R\win-library\3.6\Rio
```

While in groundhogR it is stored [^2]

[^2]: Note then, that there is a parent directory with the R version for which 
the package was installed. So there could also be a 
`C:\Users\mike\Documents\groundhogR\R-3.6.3\rio_0.5.1` and a `C:\Users\mike\Documents\groundhogR\R-3.6.1\rio_0.5.1`.

```
C:\Users\mike\Documents\groundhogR\R-3.6.2\rio_0.5.1
```

So while when with base R rio is updated, the same folder now has different
contents 

```
C:\Users\mike\Documents\R\win-library\3.6\Rio
```

With groundhogR a new folder is added, keeping both versions available:

```
C:\Users\mike\Documents\groundhogR\R-3.6.2\rio_0.5.2
C:\Users\mike\Documents\groundhogR\R-3.6.2\rio_0.5.1
```

But, when deciding whether to install a new copy of rio_0.5.1, groundhogR
checks if another identical copy already exists in an R version directory with
the same minor (e.g., R-3.6.x) and if it does, it loads it instead of
re-installing it. Unless, the `force.course = TRUE` option is checked, in which a
full R version match is enforced, and if 3.6.1 has it, but using 3.6.2, it would
be installed anyway.

## Dynamic search path

The `library()` command in R uses the `.libPaths()` path to search for installed 
packages. That path is stable for all `library()` calls (but can be modified by
users). 

`groundhog.library()`, in contrast, uses changes the searchpath for every call, 
so as to find the right version of the package for the date indicated, and R 
version being used. 

For instance, `groundhog.library("rio", "2017-10-11")` will find the version of
rio for that, date, and all the dependencies for rio for that date, and add the
path to each specific needed version to the search path, so that R uses the 
corresponding version

# Comparison to checkpoint

The checkpoint package is offered by Microsoft to enable reproducible R Code 
relying on Microsoft's MRAN archive (with daily copies of every file on CRAN, 
starting in late 2014). [^3]

[^3]: I have learned that there are several individual days missing from the 
MRAN archive, it is not literally every day.

groundhogR is thus a substitute for checkpoint.
This section discusses feature of checkpoint that were deemed undesirable and which are absent from groundhogR

1. Lack of feedback on errors to debug 

    When checkpoint fails to install a package it does not provide feedback on how
    to correct the problem. For example, the package igraph_0.7 (with over 500
    reverse dependencies), will not install in R version 3.5.0 or higher. To run a
    script that depends on it, one must use an older versions of R which will
    allow igraph 0.7 to be installed.
  
    But, the only feedback  checkpoint provides a user that attempts to install
    igraph 0.7 in a newer version of R is that the package failed to load (and
    therefore so would the 500+ packages that depend on it, or those that depend on
    those packages, etc.).
  
    groundhogR tells users which R version is needed to run igraph 0.7 (or any
    package that fails to install), and gives instructions for installing and
    running older R.

2. Multiple identical installations

    checkpoint starts an empty package library for every project, leading to new
    installations of every package needed for every project. If the same package is
    used for 20 scripts, it is installed 20 times byccheckpoint, and 20 copies are
    permanently stored.
  
    groundhogR uses a single comprehensive library that has no duplicate package
    installs.

3. Script flow

    For checkpoint to run, scripts must be kept in a project folder, and be
    already saved there.
  
    So, e.g., if a user starts a new script without saving it in an existing
    project, checkpoint will not work.  
  
    When the date for checkpoint is changed, all scripts in that project will
    reinstall and all packages for that new date (even if the new date leads to the
    same package being downloaded and installed again). From that point onwards
    users can use `install.packages()` and they will use the previously chosen date,
    but this is not self-evident from the command itself, and if a line of code is
    skipped, or a file was not saved, a different package may be installed, in ways
    that are not traceable. Mistakes seem likely.

4. No packages before 2014

    Checkpoint relies exclusively on MRAN, which started in late 2014. No earlier
    packages can be installed to render old scripts reproducible. groundhogR uses
    the CRAN archive of source files to access any CRAN package ever available.

5. Difficult to debug non-working scripts.

    If an existing script does work due to backwards incompatibility of a required
    package, and the date when the packages used in that script is unknown (a
    likely scenario), one can try to revive the script by trying alternative
    install dates for each of the packages, but this is not possible with
    checkpoint, as all packages share the same date. In addition, changing the
    date to debug one package will lead to reinstalling all packages in any script
    in that project. Thus every date attempted coud lead to several minutes of
    installation time before anything can be attempted.
